---
title: "Assignment 1"
author: "Gijs Smeets, Daan Wijnhorst, Moos Middelkoop, group 35"
date: "February 2023"
output:
  html_document:
    df_print: paged
fontsize: 11pt
highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(glue)
```

**Introductory description:** To investigate the effect of 3 types of diet, 78 persons were divided randomly in 3 groups, the first group following diet 1, second group diet 2 and the third group diet 3. Next to some other characteristics, the weight was measured before diet and after 6 weeks of diet for each person in the study. The collected data is summarized in the data frame diet.txt with the following columns: person – participant number, gender – gender (1 = male, 0 = female), age – age (years), height – height (cm), preweight – weight before the diet (kg), diet – the type of diet followed, weight6weeks – weight after 6 weeks of diet (kg). Compute and add to the data frame the variable weight.lost expressing the lost weight, to be used as response variable.

Let's start by loading the data and creating a new column containing `weight.lost`, the difference between starting weight and weight after six weeks. If participants actually lost weight, this value will be negative. If they gained weight, this value will be positive.
```{r}
data = read.table('../data/diet.txt', header=TRUE)

weight.lost = data$preweight - data$weight6weeks
data = cbind(data,weight.lost)

pre = data$preweight
post = data$weight6weeks
age = data$age
```

### A
#### Make an informative graphical summary of the data relevant for study of the effect of diet on the wight loss. 

wat is hier mis???
```{r}
par(mfrow = c(1,3))
boxplot(data$weight.lost[data$diet == 1], main="Diet 1", ylab="Weight lost")
boxplot(data$weight.lost[data$diet == 2], main="Diet 2", ylab="Weight lost")
boxplot(data$weight.lost[data$diet == 3], main="Diet 3", ylab="Weight lost")

#OR
boxplot(data$weight.lost~data$diet)
boxplot(data$weight.lost~data$gender)
hist(data$weight.lost)
```

lorem ipsum dolor sit amet

####By using only the columns preweight and weight6weeks, test the claim that the diet affects the weight loss. Check the assumptions of the test applied.

To test this claim, we use a paired t-test. Normality of data is crucial for a correct t-test. We test this assumption by creating a QQplot and performing a shapiro-wilk test. As this is a two-sample t-test, we check normality of the **differences**. If the differences follow a normal distribution, we expect the qqplot to show a linear relation and we expect the null hypothesis of normality to not be rejected by the Shapiro-Wilk test.

```{r}
qqnorm(weight.lost); qqline(weight.lost)
shapiro.test(weight.lost)
```
The results of the plot and the test are in line with our expectations, there is no reason to expect that the differences are not taken from a normal population. We can commence with a paired two-sample t-test.

```{r}
t.test(pre,post,paired=TRUE)
```
The resulting very small p value confirms the claim that the diet affects weight loss.

### B
#### Apply one-way ANOVA to test whether type of diet has an effect on the lost weight. Do all three types diets lead to weight loss? Which diet was the best for losing weight?

First we have to factorize the diet column of the data, and strip all non-interesting columns from the dataframe. We remain with a dataframe containing diet type and weight lost as columns, and do a small check to see if the diet values are now factors instead of a numeric value.

```{r}
data$diet = as.factor(data$diet)
df = data[c("weight.lost","diet")]; df
is.factor(df$diet); is.numeric(df$diet)
```

Next we can perform a one-way ANOVA. We print a summary of the results, and the resulting confidence intervals

```{r}
weightlostaov = lm(weight.lost~diet,data=df)
anova(weightlostaov)
summary(weightlostaov)
confint(weightlostaov)
```

Only diet 1 and 3 seem to have a statistically significant effect on the weight lost. Diet one looks to be the best for weight loss, with the 95% CI being [2.326, 4.274] (kilograms lost). The p value is also by far the lowest for diet 1.

We still have to show that the normality assumption is not violated, this is done by plotting the residuals. They should look normal. Furthermore, plotted against the residuals, the fitted values should show no pattern

```{r}
par(mfrow=c(1,2)); qqnorm(residuals(weightlostaov))
plot(fitted(weightlostaov),residuals(weightlostaov))
```

Judging from these plots, there is no reason to suspect the normality assumption has been violated.

#### Can the Kruskal-Wallis test be applied for this situation?

It can be used! However, we have seen that residuals are normally distributed, so there would be no need to use the Kruskal-Wallis test. Let's compare the kruskal-wallis test results with those of one-way anova for curiosity's sake.

```{r}
kruskal.test(weight.lost, df$diet)
```

### C
#### Use two-way ANOVA to investigate effect of the diet and gender (and possible interaction) on the lost weight.

First, create a new dataframe consisting of gender, diet, and weight lost. Make sure gender and diet are factors and nog numeric values.

```{r}
data$gender = as.factor(data$gender)
df2 = data[c("weight.lost","diet","gender")]
is.factor(df2$gender); is.factor(df2$diet)
```

To visually show wheter there is an interaciton effect, an interaction plot can be made. If curves run parralel, there is probably no interaction. Otherwise, interaction shows up as nonparallel curves

```{r}
gender = data$gender
diet = data$diet
par(mfrow=c(1,2))
interaction.plot(gender,diet,weight.lost)
interaction.plot(diet,gender,weight.lost)
```
From the results of these plots, we can expect an interaction effect between gender and diet. Now it's time to perform the two-way ANOVAn to statistically inspect interaction beteen gender and diet on weight loss. Remeber: when performing a two-way anova on the interaction of two factors, one cannot make conclusions based on

```{r}
twoway = lm(weight.lost~diet*gender,data=df2)
anova(twoway)
summary(twoway)
confint(twoway)
```

With p = 0.048 and alpha = 0.05 for the interaction diet*gender, the interaction variable is significant. Because we know there is an interaction effect, we can't check for individual effects reliably (KLOPT DIT????????)

In order to check the assumptions for normality and equal variances, we still need to plot the residuals.
```{r}
par(mfrow=c(1,2)); qqnorm(residuals(twoway));
plot(fitted(twoway),residuals(twoway))
```
The qqplot seems fairly linear, and there is no pattern in the residuals plotted against the fitted data. Therefore there is no reason to suspect violation of the normality and equailyt of variances assumptions.

### E
#### Which of the two approaches, the one from b) or the one from c), do you prefer? Why? For the preferred model, predict the lost weight for all three types of diet." instead of "Which of the two approaches, the one from b) or the one from d), do you prefer? Why? For the preferred model, predict the lost weight for all three types of diet for an average person.

Personally, I like the approach from b best, because it tests for an interaction effet. By my understanding, if an interaction effect between two factors is present, the results from just analyzing those factors individually, will be less reliable. This is because we know for certain that another factor had an influence on the relation between the first factor and the outcome variable.

because we already got rid of first differences (resulting in variable 'weight.lost'), we can interpret the ANOVA estimated parameters as without intercept

```{r}
summary(weightlostaov)
```
Hence, the expected weight loss for an average person in kilograms based on each three diets is: diet 1: 3.3, diet 2: 0.27 (no significant weight loss), and diet three: 1.8481.