---
title: "Assignment 1"
author: "Gijs Smeets, Daan Wijnhorst, Moos Middelkoop, group 35"
date: "February 2023"
output:
  html_document:
    df_print: paged
fontsize: 11pt
highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(glue)
```

## Exercise 1. Birthweights

LOADING IN DATA
```{r}
data = read.table('../data/birthweight.txt',header=TRUE)
x = data$birthweight
```

#### A
To check for normality we use a Shapiro-Wilk test. The null-hypothesis for this test is that the data is normally distributed. While it is not possible to infer results from a failed rejection of the null hypothesis , the result of this test can tell us that there is no significant non-normality present.

```{r}
shapiro.test(x)
```

Additionally, we create a QQ-plot to plot the quantiles of our data against those of a normal distribution, to get a visual impression of the normality of the data.

```{r,echo=FALSE,figheight=1.5}
qqnorm(x); qqline(x)
```

The linear look this plot and the result from the Shapiro-test give us a strong impression that our data is normally distributed. Now we can go on to construct the 95% confidence interval (CI) for $\mu$, assuming normality.

```{r}
mean = mean(x)
sd = sd(x)
n = length(x)

ci = c((mean - qnorm(0.98)*(sd/sqrt(n))), (mean + qnorm(0.98)*(sd/sqrt(n))))
```

This results in a confidence interval of [`r round(ci,3)`]

If we want to construct a confidence interval bounded by a length of 100, we have that the margin of error is 50. Following theory, we have to show that qnorm(1 - $\alpha$/2) * (sd / sqrt(n)) =< 50 to compute the minimal sample size needed to provide that the length of the 96%-CI is at most 100.
```{r}
min_n = ((qnorm(0.98)^2)*(sd^2))/(50^2)
```
We can draw values from the normal distribution as we have previously assumed the normality assumption of the data to be correct. This results in ``min_n` being `r round(min_n,3)`. Next we perform a bootstrap 96%-CI for mu and we compare it to the CI computed before. We take 1000 repetitions.

```{r}
B = 1000
Tstar = numeric(B);
for (i in 1:B){
  Xstar = sample(x,replace=TRUE)
  Tstar[i] = mean(Xstar)
}
Tstar20 = quantile(Tstar,0.020)[[1]]
Tstar980 = quantile(Tstar,0.975)[[1]]
bootstrapci = c(2*mean-Tstar980,2*mean-Tstar20)

lenci = round(ci[2]-ci[1],3)
lenbsci = round(bootstrapci[2]-bootstrapci[1],3)
glue('CI: [{ci[1]}, {ci[2]}],    Bootstrap CI length: [{bootstrapci[1]}, {bootstrapci[2]}]')
glue('CI length: {lenci},    Bootstrap CI length: {lenbsci}');
```

This shows that the length of the bootstrap CI is smaller and it lies inside the original CI, so we suspect that it is more accurate.

 
#### B

An expert claims that the mean birthweight is bigger than 2800 gram. In order to verify this claim, we perform a t-test.
```{r}
t.test(x, mu=2800)
```
The p-value 0.02713 < 0.05, and the mean of the sample is 2913, we can reject the null-hypothesis and conclude that the mean birthweight is statistically significantly bigger than 2800 grams.

This result is already statistically significant. One could also argue that a one-sided t-test can be justified. because it's an experts claim that babies will weigh more than 2800 grams.
This would make `H0 = mu > 2800`. 

```{r}
t.test(x, mu=2800, alt='g')
```

A one-sided test is less strict than a two-tailed test and as expected, the p value is twice as low for the one-sided test as for the two-tailed test. 

Depending on which side you are looking at, a one-sided confidence interval is either of the form (âˆ’âˆž,ð‘1] or of the form [ð‘2,âˆž)

To test about a sample median, we can also use a sign test. We propose a sign test where the H0 is median = 2800 and the H1 is median > 2800. We'll call 2800 the m0. From this follows the test statistic: T = #(x>2800). Under H0, the data should follow a binomial distribution, so we can compare a binomial distribution with the test statistic.
```{r}
nT = length(x[x>2800]); 
psign = binom.test(nT,n,p=0.5)[[3]]
psign
```

When looking at the resulting p value from the sign test, we can see that the null hypothesis doesn't get rejected, we conclude that we do not have enough evidence to reject H0 with the sample data.

### C
We test at 'some' $\mu$>2800. Here we test at $\mu$ = 2800 (we generate under H1, $\mu$ = 2800.) We can already see that the fraction of rejections is larger for t-tests (0.051 > 0.046). If we generate for $\mu$ > 2800, say, $\mu$ = 2900, then the fraction of rejections is even larger (sign test: 0.326, t-test: 0.6) than for the sign test. From this we can conclude the power of the t-test is larger than for the sign test.

```{r}
psign_val = numeric(B)
pttest_val = numeric(B)

B = 1000

for (i in 1:B) {
    x1 = rnorm(n,mean=2900,sd=sd)
    pttest_val[i]=t.test(x1,mu=2800,alt="g")[[3]]; 
    psign_val[i]=binom.test(sum(x1>2800),n,p=0.5)[[3]]; 
  }
fracsign = sum(psign_val<0.05)/B; fracsign
fracttest = sum(pttest_val<0.05)/B; fracttest
```


### D
#### Let p be the probability that birthweight of a newborn baby is less than 2600 gram. Using asymptotic normality, the expert computed the left end p-hat-l = 0.25 of the confidence interval [p-hat-l, p-hat-r] for p. Recover the whole confidence interval and its confidence level.

The best estimator we can create right now is that the  lies under 2600 grams.

```{r}
q = length(x[x<2600])/n - 0.25
pr = length(x[x<2600])/n + q
ci_p = c(0.25,pr)
```
Confidence interval is [`r round(ci_p,3)`]

Now compute confidence level, by: $q = z_(a/2) * sd/sqrt(n)$, so: $z_(a/2) = q * sqrt(n) / sd$
```{r}
z = q * (sqrt(n) / sd)
pnorm(z)
```

Is this correct? -> alpha. Look at this

### E
#### The expert also reports that there were 34 male and 28 female babies among 62 who weighted less than 2600 gram, and 61 male and 65 female babies among the remaining 126 babies. The expert claims that the mean weight is different for male and female babies. Verify this claim by an appropriate test.

To verify this claim we have to test for a difference in proportions. This can be done with `prop.test()` in R. This test tests whether the proportion of 'success' between two population differs. In order to use this test, we have to assume normality of the data. In this case we take 'succes' to mean weighing less than 2600 grams. This way we can test whether the proportion of babies weighing less than 2600 grams is different between males and females.

```{r}
# <2600g: 34 males, 28 females
# >2600g: 61 males, 65 females.
male = 34 + 61
female = 28 + 65

prop.test(c(34,28),c(male,female))
```
The results from this test show us that there is no significant difference between the proportion of babies weighing less than 2600 grams between males and females.
